Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2009-12-19 ltx/BrainIDE.tex
Initial commit of matlab scripts
Updated matlab following Parhams correction

The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
f\left( {{v_t}\left( {r'} \right)} \right) = \frac{{{\nu _0}}}{{1 + \exp \left( {\beta \left( {{v_0}\left( r \right) - {v_t}\left( {r'} \right)} \right)} \right)}}.
\end{equation}
%\begin{equation}\label{RateOfPreSynPops}
%p\left( {v\left( {r,\bar t} \right)} \right) = {\nu_0}u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right)
%\end{equation}
%\begin{equation}\label{HeavisideDef}
%u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right) = \left\{ {\begin{array}{*{20}{c}}
%   1 & {v\left( {r,\bar t} \right) \ge {v_0}\left( r \right)}  \\
%   0 & {otherwise}  \\
%\end{array}} \right.
%\end{equation}
Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining equations~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega  h\left(t - t'\right)w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' dt'+e\left(r,t\right)
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {r,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  +e\left( {r,t} \right).
\end{equation}
\subsection{Multi-Layer Model}
\begin{equation}\label{ExcitatoryLayer}
\frac{dv_e\left( {r,t} \right)}{{dt}} + \alpha_ev_e\left( {r,t} \right) = \int_\Omega  {w_{ee}\left( {r,r'} \right)f\left( v_e\left( r',\bar t \right) \right) + w_{ei}\left( {r,r'} \right)f\left( v_i\left( r',\bar t \right) \right)dr'}  +\varepsilon_e\left( {r,t} \right)
\end{equation}
\begin{equation}\label{InhibitoryLayer}
\frac{dv_i\left( {r,t} \right)}{{dt}} + \alpha_iv_i\left( {r,t} \right) = \int_\Omega  {w_{ie}\left( {r,r'} \right)f\left( v_e\left( r',\bar t \right) \right) + w_{ii}\left( {r,r'} \right)f\left( v_i\left( r',\bar t \right) \right)dr'}  +\varepsilon_i\left( {r,t} \right)
\end{equation}
The subscript $ei$ denotes connectivity from an inhibitory neural element to an excitatory neural element. We can write the model in matrix form
\begin{equation}\label{matrixEquation}
\left[ {\begin{array}{*{20}{c}}
   {\frac{{d{v_e}\left( t \right)}}
{{dt}}}  \\
   {\frac{{d{v_i}\left( t \right)}}
{{dt}}}  \\

@@

 \end{array} } \right]} \left[ {\begin{array}{*{20}{c}}
   {f\left( {{v_e}\left( {r',t} \right)} \right)}  \\
   {f\left( {{v_i}\left( {r',t} \right)} \right)}  \\

@@

 \end{array} } \right]
\end{equation}
The iEEG observation equation is
\begin{equation}\label{iEEGObservationEquation}
    y(k,t) = \int_{\Omega} m(k,r)\left(v_e(r,t) + v_i(r,t)\right)dr' + w(k,t) 
\end{equation}

%\subsection{Alternate Model Description I}
%An alternate and equivalent way to remove the temporal integration using the Laplace transform . We start by considering a discrete neural mass $i$ with inputs with zero delay from another discrete mass $j$ and
%\begin{equation}\label{TwoDiscreteMasses}
%{v_i}\left( t \right) = \int_{ - \infty }^t {{h_j}(t - t'){w_{ij}}p\left( {{v_j}\left( t \right)} \right)dt'},
%where $w_{ij}$ can be considered the synaptic efficiency from $j$ to $i$. Now we take the Laplace transform $\mathcal{L}(v_i(t))=V(s)$ yielding
%  {V_i}\left( s \right) &=& {H_j}\left( s \right){w_{ij}}P\left( {{V_j}\left( s \right)} \right) \nonumber \\
%   &=& {\left( {s + \alpha} \right)^{ - 1}}{w_{ij}}P\left( {{V_j}\left( s \right)} \right)  \\
%  \Rightarrow s{V_i}\left( s \right) + \alpha{V_i}\left( s \right) &=& {w_{ij}}P\left( {{V_j}\left( s \right)} \right).
%Now taking the inverse Laplace transform we get
%\begin{equation}\label{InverseLaplace}
%\frac{{d{v_i}\left( t \right)}}{{dt}} + \alpha{v_i}\left( t \right) = {w_{ij}}p\left( {{v_j}\left( t \right)} \right).
%Next we convert the model from discrete neural masses to a continuum by
%\begin{equation}\label{Discrete2Continuum}
%\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)p\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {s,t} \right).
%\subsection{Alternate Model Description II}
%\begin{figure*}
%\centering
%\includegraphics[scale=0.4]{NeuralMassFigure.eps}
%\caption[NeuralMassModel]{This figure describes the basic interactions of neural masses that constitute the neural field model.}
%\label{NeuralMassModel}
%\end{figure*}
%An alternative description of the model is as follows. The model can be thought of as a continuum of neural masses. To illustrate this idea, we shall first describe the interaction of discrete neural masses as seen in Figure~\ref{NeuralMassModel}. Assuming that post-synaptic potentials sum linearly, the membrane potential at the soma of neuron $i$ is described as
%\begin{equation}\label{VoltageFromPSPs}
%{v_i}\left( t \right) = \sum\nolimits_{j,k} {{w_{ij}}{h_{j}}\left( t - {t_k}\right)} u\left( {t - {t_k}} \right),
%where $w_{ij}$ is the synaptic weight from neuron $j$ to $i$, $h(t)$ is the synaptic response kernel, $S$ is the Heaviside step function and action potential arrive at neuron $i$ at times $t_k$. Alternatively the membrane potential can described by the convolution of synaptic response kernel with the pre-synaptic firing rate $r_j(t)$ where
%\begin{equation}\label{VoltageFromRate}
%{v_i}\left( t \right) = \sum\nolimits_j {\int_{ - \infty }^t {{w_{ij}}{h_j}\left( {t - t'} \right){r_j}\left( t \right)} } dt'.
%To convert the equation system into an integral differential system, we shall take derivative of~\ref{EqVoltageFromPSPs} yielding
%  \frac{{d{v_i}\left( t \right)}}{{dt}} &=& \sum\nolimits_{j,k} {\left( {{w_{ij}}{h_j}\left( t- {t_k} \right)\frac{d}{{dt}}u\left( {t - {t_k}} \right) + {w_{ij}}u\left( {t - {t_k}} \right)\frac{d}{{dt}}{h_j}\left( t- {t_k} \right)} \right)} \nonumber \\
%   &=& \sum\nolimits_{j,k} {{w_{ij}}{h_j}\left( t- {t_k} \right)\delta \left( {t - {t_k}} \right)}  - \sum\nolimits_{j,k} {{w_{ij}}u\left( {t - {t_k}} \right)\alpha_j{h_j}\left( t- {t_k} \right)}   \nonumber \\
%   &=& \sum\nolimits_{j,k} {{w_{ij}}\delta \left( {t - {t_k}} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j} \nonumber  \\
%   &=& \sum\nolimits_j {{w_{ij}}{\nu_j}\left( t \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j} \nonumber \\
%   &=& \sum\nolimits_j {{w_{ij}}p\left( {{v_j}\left( t \right)} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j}.
%This form can easily be extended to represent a continuous field.

@@

To infer the state from observations we can write the likelihood function
\begin{equation}\label{Likelihood}
p\left( {{y_{1,...,T}}|{v_{1,...,T}}} \right) = \prod\limits_{t = 1}^T {p\left( {{y_t}|{v_t}} \right)} \prod\limits_{t = 2}^T {p\left( {{v_{t + 1}}|{v_t}} \right)} p\left( {{v_0}} \right)
\end{equation}
where
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  p\left( {{y_t}|{v_t}} \right) &=& \mathcal{N}\left( {\int_\Omega  {m\left( {k,r} \right)v_t( r )dr} ,R} \right) \\
  p\left( {{v_{t + 1}}|{v_t}} \right) &=& ??? \\
  p\left( {{v_0}} \right) &=& \mathcal{GP}\left( {{\mu _0},{\pi _0}} \right).
\end{eqnarray}
We can not calculate the posterior in this Bayesian model for some reason (Mike and Ken can elaborate), so we need to decompose the field into weighted basis functions.

>>>>

The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
f\left( {{v_t}\left( {r'} \right)} \right) = \frac{{{\nu _0}}}{{1 + \exp \left( {\beta \left( {{v_0}\left( r \right) - {v}\left(r',\bar{t} \right)} \right)} \right)}}.
\end{equation}
%\begin{equation}\label{RateOfPreSynPops}
%p\left( {v\left( {r,\bar t} \right)} \right) = {\nu_0}u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right)
%\end{equation}
%\begin{equation}\label{HeavisideDef}
%u\left( {v\left( {r,\bar t} \right) - {v_0}\left( r \right)} \right) = \left\{ {\begin{array}{*{20}{c}}
%   1 & {v\left( {r,\bar t} \right) \ge {v_0}\left( r \right)}  \\
%   0 & {otherwise}  \\
%\end{array}} \right.
%\end{equation}
Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining equations~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t\int\limits_\Omega  h\left(t - t'\right)w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' dt'+e\left(r,t\right)
\end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. This provides the most general form
\begin{equation}\label{GenForm}
L\left( {\frac{\partial }{{dt}}} \right)v\left( {r,t} \right) = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {r,t} \right),
\end{equation}
where $L$ is a polynomial of order $n$ with constant coefficients that provides an $n^{th}$ order model. Most studies considering this model use first or second order time derivatives. We shall use the first order model giving
\begin{equation}\label{FinalForm1}
\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)f\left( {v\left( {r',\bar t} \right)} \right)dr'}  +e\left( {r,t} \right).
\end{equation}
\subsection{Multi-Layer Model}
\begin{equation}\label{ExcitatoryLayer}
\frac{dv_e\left( {r,t} \right)}{{dt}} + \alpha_ev_e\left( {r,t} \right) = \int_\Omega  {w_{ee}\left( {r,r'} \right)f\left( v_e\left( r',\bar t \right) \right) + w_{ei}\left( {r,r'} \right)f\left( v_i\left( r',\bar t \right) \right)dr'}  +\varepsilon_e\left( {r,t} \right)
\end{equation}
\begin{equation}\label{InhibitoryLayer}
\frac{dv_i\left( {r,t} \right)}{{dt}} + \alpha_iv_i\left( {r,t} \right) = \int_\Omega  {w_{ie}\left( {r,r'} \right)f\left( v_e\left( r',\bar t \right) \right) + w_{ii}\left( {r,r'} \right)f\left( v_i\left( r',\bar t \right) \right)dr'}  +\varepsilon_i\left( {r,t} \right)
\end{equation}
The subscript $ei$ denotes connectivity from an inhibitory neural element to an excitatory neural element. We can write the model in matrix form
\begin{equation}\label{matrixEquation}
\left[ {\begin{array}{*{20}{c}}
   {\frac{{d{v_e}\left( t \right)}}
{{dt}}}  \\
   {\frac{{d{v_i}\left( t \right)}}
{{dt}}}  \\

@@

 \end{array} } \right]} \left[ {\begin{array}{*{20}{c}}
   {f\left( {{v_e}\left( {r',\bar{t}} \right)} \right)}  \\
   {f\left( {{v_i}\left( {r',\bar{t}} \right)} \right)}  \\

@@

 \end{array} } \right]
\end{equation}

\begin{equation}\label{matrix_equation_in_vector_notation}
   \frac{d\mathbf{V}(r,t)}{dt} = \int_\Omega \mathbf{W}(r,r')f\left(\mathbf{V}(r,t)\right)dr' - \mathbf{A}\mathbf{V}(r,t) + \mathbf{E}(r,t)
\end{equation}

The iEEG observation equation is
\begin{equation}\label{iEEGObservationEquation}
    y(k,t) = \int_{\Omega} m(k,r)\left(v_e(r,t) + v_i(r,t)\right)dr' + w(k,t)
\end{equation}

\subsection{Alternate Model Description I}
An alternate and equivalent way to remove the temporal integration using the Laplace transform . We start by considering a discrete neural mass $i$ with inputs with zero delay from another discrete mass $j$ and
\begin{equation}\label{TwoDiscreteMasses}
{v_i}\left( t \right) = \int_{ - \infty }^t {{h_j}(t - t'){w_{ij}}p\left( {{v_j}\left( t \right)} \right)dt'},
\end{equation}
where $w_{ij}$ can be considered the synaptic efficiency from $j$ to $i$. Now we take the Laplace transform $\mathcal{L}(v_i(t))=V(s)$ yielding
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  {V_i}\left( s \right) &=& {H_j}\left( s \right){w_{ij}}P\left( {{V_j}\left( s \right)} \right) \nonumber \\
   &=& {\left( {s + \alpha} \right)^{ - 1}}{w_{ij}}P\left( {{V_j}\left( s \right)} \right)  \\
  \Rightarrow s{V_i}\left( s \right) + \alpha{V_i}\left( s \right) &=& {w_{ij}}P\left( {{V_j}\left( s \right)} \right).
\end{eqnarray}
Now taking the inverse Laplace transform we get
\begin{equation}\label{InverseLaplace}
\frac{{d{v_i}\left( t \right)}}{{dt}} + \alpha{v_i}\left( t \right) = {w_{ij}}p\left( {{v_j}\left( t \right)} \right).
\end{equation}
Next we convert the model from discrete neural masses to a continuum by
\begin{equation}\label{Discrete2Continuum}
\frac{{dv\left( {r,t} \right)}}{{dt}} + \alpha{v\left( {r,t} \right)} = \int_\Omega  {w\left( {r,r'} \right)p\left( {v\left( {r',\bar t} \right)} \right)dr'}  + e\left( {s,t} \right).
\end{equation}

%\subsection{Alternate Model Description II}
%\begin{figure*}
%\centering
%\includegraphics[scale=0.4]{NeuralMassFigure.eps}
%\caption[NeuralMassModel]{This figure describes the basic interactions of neural masses that constitute the neural field model.}
%\label{NeuralMassModel}
%\end{figure*}
%An alternative description of the model is as follows. The model can be thought of as a continuum of neural masses. To illustrate this idea, we shall first describe the interaction of discrete neural masses as seen in Figure~\ref{NeuralMassModel}. Assuming that post-synaptic potentials sum linearly, the membrane potential at the soma of neuron $i$ is described as
%\begin{equation}\label{VoltageFromPSPs}
%{v_i}\left( t \right) = \sum\nolimits_{j,k} {{w_{ij}}{h_{j}}\left( t - {t_k}\right)} u\left( {t - {t_k}} \right),
%\end{equation}
%where $w_{ij}$ is the synaptic weight from neuron $j$ to $i$, $h(t)$ is the synaptic response kernel, $S$ is the Heaviside step function and action potential arrive at neuron $i$ at times $t_k$. Alternatively the membrane potential can described by the convolution of synaptic response kernel with the pre-synaptic firing rate $r_j(t)$ where
%\begin{equation}\label{VoltageFromRate}
%{v_i}\left( t \right) = \sum\nolimits_j {\int_{ - \infty }^t {{w_{ij}}{h_j}\left( {t - t'} \right){r_j}\left( t \right)} } dt'.
%\end{equation}
%To convert the equation system into an integral differential system, we shall take derivative of~\ref{EqVoltageFromPSPs} yielding
%\begin{eqnarray}
%% \nonumber to remove numbering (before each equation)
%  \frac{{d{v_i}\left( t \right)}}{{dt}} &=& \sum\nolimits_{j,k} {\left( {{w_{ij}}{h_j}\left( t- {t_k} \right)\frac{d}{{dt}}u\left( {t - {t_k}} \right) + {w_{ij}}u\left( {t - {t_k}} \right)\frac{d}{{dt}}{h_j}\left( t- {t_k} \right)} \right)} \nonumber \\
%   &=& \sum\nolimits_{j,k} {{w_{ij}}{h_j}\left( t- {t_k} \right)\delta \left( {t - {t_k}} \right)}  - \sum\nolimits_{j,k} {{w_{ij}}u\left( {t - {t_k}} \right)\alpha_j{h_j}\left( t- {t_k} \right)}   \nonumber \\
%   &=& \sum\nolimits_{j,k} {{w_{ij}}\delta \left( {t - {t_k}} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j} \nonumber  \\
%   &=& \sum\nolimits_j {{w_{ij}}{\nu_j}\left( t \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j} \nonumber \\
%   &=& \sum\nolimits_j {{w_{ij}}p\left( {{v_j}\left( t \right)} \right)}  - {v_i}\left( t \right)\sum\nolimits_j {\alpha_j}.
%\end{eqnarray}
%This form can easily be extended to represent a continuous field.

@@

%To infer the state from observations we can write the likelihood function
%\begin{equation}\label{Likelihood}
%p\left( {{y_{1,...,T}}|{v_{1,...,T}}} \right) = \prod\limits_{t = 1}^T {p\left( {{y_t}|{v_t}} \right)} \prod\limits_{t = 2}^T {p\left( {{v_{t + 1}}|{v_t}} \right)} p\left( {{v_0}} \right)
%\end{equation}
%where
%\begin{eqnarray}
%% \nonumber to remove numbering (before each equation)
%  p\left( {{y_t}|{v_t}} \right) &=& \mathcal{N}\left( {\int_\Omega  {m\left( {k,r} \right)v_t( r )dr} ,R} \right) \\
%  p\left( {{v_{t + 1}}|{v_t}} \right) &=& ??? \\
%  p\left( {{v_0}} \right) &=& \mathcal{GP}\left( {{\mu _0},{\pi _0}} \right).
%\end{eqnarray}
%We can not calculate the posterior in this Bayesian model for some reason (Mike and Ken can elaborate), so we need to decompose the field into weighted basis functions.
